---
title: "DC Crimes"
author: "Group 5"
date: "11/11/2021"
output: html_notebook
---

### Loading Packages

For our analysis, in addition to the state-wise data package, we will use the following packages for data wrangling and visualization.

- **tidyverse** for data wrangling and visualization
- **readr** for reading csv file locally
- **lubridate** package for handling dates
- **sf** to convert coordinates into geometrical points
- **here** to specify where files/data is in the computer relative to a particular file
- **tmap** to visualize spatial data distributions by creating geographical maps by variables
- **janitor** for examining and cleaning untidy data

```{r}
library(tidyverse)
library(readr)
library(lubridate)
library(sf)
library(here)
library(tmap)
library(janitor)
```

### Loading Data

The data comes from a data set posted by Open Data DC from their own website that consists of all the crime data for past years. The data can be downloaded in .csv, .geojson and various other formats. The data is available at (“https://opendata.dc.gov/datasets/crime-incidents-in-2021/explore”) and is updated daily.

```{r}
data <- read_csv("~/DS 241/DC Crime grp/final.csv")

neigh <- st_read(here("data", "dc_neigh.geojson")) %>% 
  clean_names()
```

### Prepping Data

The data set was very untidy and `clean_names()` couldn't solve all of the issues so I tried to prep the data in a way that would be suitable for me to use. I edited the data set.
`rename()` was used to rename certain variables.
`substr()` was used to edit the data values i.e. to remove certain number of characters from either side.
`as_datetime()` helps un convert date-time values in a proper date-time format so that we ca use `lubridate` to make visualizations.

```{r}
main <- data %>%
  select(CCN:OBJECTID) %>%
  rename(complaint_num = CCN) %>%
  rename(report_date = REPORT_DAT) %>%
  rename(shift = SHIFT) %>%
  rename(method = METHOD) %>%
  rename(offense = OFFENSE) %>%
  rename(block = BLOCK) %>%
  rename(x_block = XBLOCK) %>%
  rename(y_block = YBLOCK) %>%
  rename(ward = WARD) %>%
  rename(anc = ANC) %>% # Advisory Neighborhood Commission
  rename(district = DISTRICT) %>%
  rename(psa = PSA) %>% # Police Service Area
  rename(nbhd_cluster = NEIGHBORHOOD_CLUSTER) %>%
  rename(block_grp = BLOCK_GROUP) %>%
  rename(census_tract = CENSUS_TRACT) %>%
  rename(precinct = VOTING_PRECINCT) %>%
  rename(latitude = LATITUDE) %>%
  rename(longitude = LONGITUDE) %>%
  rename(bid = BID) %>% # Business Improvement District
  rename(start = START_DATE) %>%
  rename(end = END_DATE) %>%
  rename(object_id = OBJECTID) %>%
  mutate(report_date = substr(report_date, 1, nchar(report_date)-3),
         report_date = as_datetime(report_date),
         start = substr(start, 1, nchar(start)-3),
         start = as_datetime(start),
         end = substr(end, 1, nchar(end)-3),
         end = as_datetime(end),
         nbhd_cluster = as.numeric(substring(nbhd_cluster, 9)),
         precinct = as.numeric(substring(precinct, 10)),
         block_grp = as.numeric(substring(block_grp, 8)),
         census_tract = as.numeric(census_tract),
         shift=str_to_sentence(shift),
         method=str_to_sentence(method),
         offense=str_to_sentence(offense),
         bid=str_to_title(bid)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
```

### Prepping Spatial data

```{r}
df = st_join(main, neigh)
```

### T-map for crimes in neighborhood

The tmap code below has been set to view mode. Each neighborhood has been represented by a a different shade and been filled with the area.The colored dots represents any crime that has happened in that particular area with each color representing a different type of crime.

```{r}
tmap_mode("view")

c <- tm_shape(neigh) + 
  tm_polygons(col="shapearea", palette="Blues") +
tm_shape(main) + 
  tm_dots(col="offense", palette = "Set1", stretch.palette = FALSE, size = 0.02, shape = 2) +  
  tm_layout(legend.outside = TRUE) 

c
```

### Saving t_map

```{r}
tmap_save(c, here("output", "crime.html"))
```

### Making Visualizations from the Data

The crime data from Open Data DC is from Jan 1, 2021 till Dec 8, 2021. We used some of the variables from the data set to create plots that could give us certain visualizations about crime happening in DC.

#### Time of the day

The ggplot shows us visualizations for crime by time of day.

The graph shows us that the most crimes happened during evening time. If we look at the hourly data for the number of crimes, we can see that during day time there are very less crimes compared to that in the evenings. 5:00 pm - 6:00 pm seem to have most crimes reported. We can see a trend in the graph below, as the day gets brighter the crimes decrease a lot and start increasing as the day goes by.

```{r}
ggplot(main, aes(factor(hour(report_date)))) +
  geom_bar() +
  labs(
    x="Year",
    y="Number of Crimes Reported",
    title="Total Number of crimes by hour") +
    theme(legend.position = "none")
```
The `ggsave()` code has been used with the `here` package. `ggsave()` ensures that the previous plot created gets saved with the name and device as declared bu the user. In this case, it will save it in the file/folder as instructed by the user. I will be saving all my necessary plots in the output folder.

```{r}
ggsave(here("output", "hourly_crime.jpg"))
```

#### Method

The ggplot shows us visualizations for crime by method where "other" form of weapon was mostly used where "other" could be physically, or any other means of attack.

```{r}
ggplot(main, aes(x=fct_infreq(method), fill=method)) +
  geom_bar() +
  labs(
    x="Method of Crime",
    y="Number of Crimes Reported",
    fill="Method") +
  scale_y_continuous(labels = scales::comma)+
  theme(legend.position = "none")
```

#### Offense

`ggplot` code below shows the number of crimes by offense. In the plot below, we see that there are no records for Arson related crimes. However, that is because the crimes related to Arson very less (3) compared to that of Theft/Other (10236).

```{r}
ggplot(main, aes(x=fct_infreq(offense), fill=offense)) +
  geom_bar(stat='count') +
  coord_cartesian(ylim=c(0,65000))+
  labs(
    x="Criminal Offense",
    y="Number of Crimes Reported",
    fill="Offense") +
    theme(axis.text.x = element_text(angle = 30, hjust=1)) +
    theme(legend.position = "none")
```

#### Total Crimes since 2017 (by month)

```{r}
ggplot(main, aes(factor(month(report_date, label=TRUE)))) +
  geom_bar() +
  labs(
    x="Year",
    y="Number of Crimes Reported") +
    theme(legend.position = "none")
```

#### Yearwise Number of crimes

If we see the data carefully, we will notice that the number of crimes in the years 2017 to 2018 to 2019 increased at a very constant rate. However, for 2020 the number went down by nearly 5000 cases, which is obviously due to COVID-19. Lesser people came out of their houses that resulted in lesser crimes being committed. Even for 2021, the data is until mid-November and the number seems fairly low due to COVID-19.

```{r}
ggplot(main, aes(factor(year(report_date)))) +
  geom_bar() +
  labs(
    x="Year",
    y="Number of Crimes Reported") +
    theme(legend.position = "none")
```

#### Did COVID impact crimes in December 2019 ?

We can see that the number of crimes in November and December of 2019 went down compared to that of October. We can see in the graph that the number of crimes was increasing constantly over each month but failed to increase for Nov/Dec because that was when COVID started spreading and was possibly one of the reasons why the numbers went down.

```{r}
data_2019 <- main %>%
  filter(year(report_date)==2019)
  
ggplot(data_2019, aes(factor(month(report_date, label=TRUE)))) +
  geom_bar() +
  labs(
    x="Year",
    y="Number of Crimes Reported",
    title="Number of Crimes Reported per month in 2019") +
    theme(legend.position = "none")
```

#### How about 2020 ?

The second wave of COVID-19 hit the United States around the months of April-May, and we can clearly see the the number of crimes reported went down during those months as there was a strict lock down going on.

```{r}
data_2020 <- main %>%
  filter(year(report_date)==2020)

ggplot(data_2020, aes(factor(month(report_date, label=TRUE)))) +
  geom_bar() +
  labs(
    x="Year",
    y="Number of Crimes Reported",
    title="Number of Crimes Reported per month in 2020",
    subtitle="The second wave of COVID hit around April-May, and the number of crimes reported went down 
significantly.") +
  theme(legend.position = "none")
``` 

The `ggsave()` code has been used with the `here` package. `ggsave()` ensures that the previous plot created gets saved with the name and device as declared bu the user. In this case, it will save it in the file/folder as instructed by the user. I will be saving all my necessary plots in the output folder.

```{r}
ggsave(here("output", "crime_2020.jpg"))
```