---
title: "Lab 04 - La Quinta is Spanish for next to Denny's, Pt. 1"
author: "Shreejit Poudyal"
date: "10/08/2021"
output: html_notebook
---

### Load packages and data

For the analysis, I used the following packages for data wrangling and visualisation.

- **tidyverse** for data wrangling and visualization
- **mapview** to create interactive maps
- **dsbox** package will provide us with all the necessary data for DC Bikeshare.
- **readr** to read large flat files quickly
- **here** to specify where files/data is in the computer relative to a particular file

```{r load-packages, message = FALSE, warning=FALSE}
library(tidyverse)
library(mapview)
library(dsbox)  
library(readr)
library(here)
```

### Load Data

The exercise can be found here: https://rstudio-education.github.io/datascience-box/course-materials/lab-instructions/lab-04/lab-04-viz-sp-data.html

```{r load-data, message = FALSE, warning=FALSE}
states <- read_csv("https://raw.githubusercontent.com/rstudio-education/datascience-box/master/course-materials/lab-instructions/lab-04/data/states.csv")
```

### Exercise 1

What are the dimensions of the Denny’s dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the data set represent? What are the variables?

The following code gives us the dimensions of the data set `dennys`. There are 1643 rows and 6 columns (originally). The rows represent the data values and the columns are variables. The variable names are address, city, state, zip, longitude and latitude.

``` {r}
dim(dennys)
glimpse(dennys)
view(dennys)
colnames(dennys)
```

### Exercise 2

What are the dimensions of the La Quinta’s dataset? What does each row in the dataset represent? What are the variables?

The following code gives us the dimensions of the data set `laquinta`. There are 909 rows and 6 columns (originally). The rows represent the data values and the columns are variables. The variable names are address, city, state, zip, longitude and latitude.

``` {r}
dim(laquinta)
glimpse(laquinta)
view(laquinta)
colnames(laquinta)
```

### Exercise 3

Take a look at the websites that the data come from (linked above). Are there any La Quinta’s locations outside of the US? If so, which countries? What about Denny’s?

There are no locations outside the United States for Dennys but there are 4 outside countries for Laquinta that are Canada, Mexico, Honduras and Colombia. However, according to the Laquinta website, they now have Laquinta located in other countries like New Zealand, Chile,Turkey and UAE.

### Exercise 4

Now take a look at the data. What would be some ways of determining whether or not either establishment has any locations outside the US using just the data (and not the websites). Don’t worry about whether you know how to implement this, just brainstorm some ideas. Write down at least one as your answer, but you’re welcomed to write down a few options too.

1.  We could create two data tables with just the state names (abbreviated form) and compare the two data tables with `states.csv` file to see which state abbreviations are invalid (meaning that those abbreviations are for states/cities outside the US.)
2.  We could create such a table that combines the values of the data tables in a way that it filters US states and Outside States by comparing the data from each individual sata sets.

### Exercise 5

Find the Denny’s locations that are outside the US, if any. To do so, filter the Denny’s locations for observations where `state` is not in `states$abbreviation`. The code for this is given below. Note that the `%in%` operator matches the states listed in the state variable to those listed in `states$abbreviation`. The `!` operator means not. Are there any Denny’s locations outside the US?

As the data shows, there are no outside US location for `dennys`.

``` {r}
dennys %>%
  filter(!(state %in% states$abbreviation))
```

### Exercise 6

Add a country variable to the Denny’s dataset and set all observations equal to "United States". Remember, you can use the mutate function for adding a variable. Make sure to save the result of this as dennys again so that the stored data frame contains the new variable going forward.

``` {r}
dennys %>%
  mutate(country = "United States")
```

### Exercise 7

Find the La Quinta locations that are outside the US, and figure out which country they are in. This might require some googling. Take notes, you will need to use this information in the next exercise.

As the data shows, there are 14 different observations that are not from the US. La Quinta locations for for abroad nations are: Canada, Mexico, New Zealand, Honduras, Turkey, United Arab Emirates, Chile and Colombia.

``` {r}
laquinta %>%
  filter(!(state %in% states$abbreviation))
dennys %>%
  filter(!(state %in% states$abbreviation))
```

### Exercise 8

Add a country variable to the La Quinta dataset. Use the case_when function to populate this variable. You’ll need to refer to your notes from Exercise 7 about which country the non-US locations are in. Here is some starter code to get you going:

In this chunk, I have added the different countries into the data set as there are various locations that are outside the US.

``` {r}
laquinta <- laquinta %>%
  mutate(laquinta, country = case_when(
    state %in% state.abb  ~ "United States",
    state %in% c("ON", "BC") ~ "Canada",
    state == "ANT" ~ "Colombia",
    state %in% c("AG", "QR", "CH", "NL", "VE", "PU", "SL") ~ "Mexico",
    state == "FM" ~ "Honduras"
  ))
dennys <- dennys %>%
  mutate(dennys, country = case_when(
    state %in% state.abb  ~ "United States",
  state=="DC" ~ "United States"
  ))
```

### Exercise 9

Which states have the most and fewest Denny’s locations? What about La Quinta? Is this surprising? Why or why not?

For `laquinta`, states like Arkansas, Iowa, Maine, Michigan, North Hampshire, Rhode Island, South Dakota, West Virginia and Wyoming have 3 or less than 3 La Quinta locations, possibly because these states aren't very densely populated.

``` {r}
dennys %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))
laquinta %>%
  count(state) %>%
  inner_join(states, by = c("state" = "abbreviation"))
```


### Exercise 10

Which states have the most Denny’s locations per thousand square miles? What about La Quinta?

``` {r}
dennys <- dennys %>%
  mutate(establishment = "Denny's")
laquinta <- laquinta %>%
  mutate(establishment = "La Quinta")
dn_lq <- bind_rows(dennys, laquinta)
ggplot(dn_lq, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point(alpha=0.5) +
  labs (
    color = "Establishment",
    title="Plotted locations for Denny's and Laquinta"
  )
mapview(dn_lq, xcol = "longitude", ycol = "latitude", zcol="establishment", crs = 4269, grid = FALSE)
```
The `ggsave()` code has been used with the `here` package. `ggsave()` ensures that the previous plot created gets saved with the name and device as declared bu the user. In this case, it will save it in the file/folder as instructed by the user. I will be saving all my necessary plots in the output folder.

```{r}
ggsave(here("output", "dennys_laquinta.jpg"))
```

### Exercise 11

Filter the data for observations in North Carolina only, and recreate the plot. You should also adjust the transparency of the points, by setting the alpha level, so that it’s easier to see the over plotted ones. Visually, does Mitch Hedberg’s joke appear to hold here?

While it can be seen that some places have both Denny's and La Quinta, there are several installments of these places that are not next to one another, so I think Mitch Hedberg's joke is true only true to a certain extent. However, if we zoom into the map, we can see that these 2 locations are at least 2-3 blocks away from one another and not "right next to each other".


``` {r}
dennys_nc = dennys %>%
  filter(state=="NC")%>%
  mutate(establishment = "Denny's")
laquinta_nc = laquinta %>%
  filter(state=="NC")%>%
  mutate(establishment = "La Quinta")
dn_lq_nc <- bind_rows(dennys_nc, laquinta_nc)

ggplot(dn_lq_nc, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point(alpha=0.5)

mapview(dn_lq_nc, xcol = "longitude", ycol = "latitude", zcol="establishment", crs = 4269, grid = FALSE)
```

### Exercise 12

Now filter the data for observations in Texas only, and recreate the plot. Visually, does Mitch Hedberg’s joke appear to hold here?

The same thing applies for Texas as well, for places that are densely populated or might have more visits to dining locations and hotels, it is obvious that they might be right next to one another. However, if we zoom into the map, we can see that these 2 locations are at least 2-3 blocks away from one another and not "right next to each other".

``` {r}
dennys_tx = dennys %>%
  filter(state=="TX")%>%
  mutate(establishment = "Denny's")
laquinta_tx = laquinta %>%
  filter(state=="TX")%>%
  mutate(establishment = "La Quinta")
dn_lq_tx <- bind_rows(dennys_tx, laquinta_tx)

ggplot(dn_lq_tx, mapping = aes(x = longitude, y = latitude, color = establishment)) +
  geom_point(alpha=0.5)

mapview(dn_lq_tx, xcol = "longitude", ycol = "latitude", zcol="establishment", crs = 4269, grid = FALSE)
```